{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2025-11-19 mssm.tex
Fixes for elcture 20.11.25

Some  examples of other $\beta$-functions in the MSSM:
\begin{itemize}
\item For the supersymmetry parameter $\mu$
\begin{equation}
\beta_{\mu}\equiv \frac{d}{dt} \mu = \frac{1}{16\pi^2}\mu [3|y_t|^2+3|y_b|^2+|y_\tau|^2-3g_2^2-\frac{3}{5}g_1^2],
\label{eq:mu_beta}
\end{equation}
where $y_t$, $y_b$ and $y_\tau$, are the top and bottom quark, and tau lepton Yukawa couplings. Other Yukawa couplings also contribute to one-loop order, but due to their small size they can usually be safely ignored. Since the gauge couplings appear on the right-hand side here, we see that the differential equations that determine the running of $mu$ and the gauge couplings are coupled. It is a general property of the $\beta$-functions that they couple the running of all parameters, which is caused by the (eventual) appearance of all particles as virtual particles in loops when we go to high-enough order.
\item For the soft-breaking parameters $M_i$ in (\ref{eq:gensoftterms}) we get the one-loop $\beta$-functions
\begin{equation}
\beta_{M_i}\equiv \frac{d}{dt}M_i = \frac{1}{8\pi^2}g_i^2M_ib_i,
\label{eq:beta_Mi}
\end{equation}
where $b_i$ is given above.
\item For the Higgs soft masses
\begin{eqnarray}
\beta_{m_{H_u}^2} &\equiv& \frac{dm_{H_u}^2}{dt} = \frac{1}{16\pi^2}(X_t-\frac{6}{5} g_1^2|M_1|^2- 6g_2^2|M_2|^2)  ,\label{eq:beta_Hu2}\\
\beta_{m_{H_d}^2} &\equiv& \frac{dm_{H_d}^2}{dt} = \frac{1}{16\pi^2}( X_b +X_\tau -\frac{6}{5} g_1^2|M_1|^2 - 6g_2^2|M_2|^2) \label{eq:beta_Hd2},
\end{eqnarray}
where we have used the standard abbreviations
\begin{eqnarray*}
X_t &\equiv& 6 |y_t|^2(m_{H_u}^2 + m_{Q_3}^2 + m_{u_3}^2) + 6|a_t|^2,\\
X_b &\equiv& 6 |y_b|^2(m_{H_d}^2 + m_{Q_3}^2 + m_{d_3}^2) + 6|a_b|^2,\\
X_\tau &\equiv& 2 |y_\tau|^2(m_{H_d}^2 + m_{L_3}^2 + m_{e_3}^2) + 2|a_\tau|^2,
\end{eqnarray*}
and where $m_{Q_3} = m_{33}^Q$, $m_{u_3}  = m_{33}^u$, $m_{d_3} = m_{33}^d$, $a_t=a_{33}^u$, $a_b=a_{33}^d$, and $a_\tau=a_{33}^e$,  in a simplification of our previous notation for the soft-term parameters in Sec.~\ref{sec:MSSM_soft_terms}.  
\end{itemize}

@@

For the scalar Higgs component fields in the MSSM this gives the potential
\begin{eqnarray}\label{eq:scaH}
V(H_u,H_d) &=& |\mu|^2 (|H_u^0|^2 + |H_u^+|^2 + |H_d^0|^2 + |H_d^-|^2)  \text{\indent \indent\indent \indent\indent\indent\indent}\text{(from $F$-terms)} \nonumber\\
&&+\frac{1}{8} (g^2 + g'{}^2)(|H^0_u|^2 + |H_u^+|^2 - |H^0_d|^2 - |H_d^-|^2)^2  \text{\indent\indent\indent\indent(from $D$-terms)} \nonumber\\
&&+ \frac{1}{2}g^2|H^+_uH_d^0{}^* + H^0_u H^-_d{}^*|^2 \nonumber\\
&&+ m_{H_u}^2(|H_u^0|^2 + |H_u^+|^2) + m_{H_d}^2(|H_d^0|^2 + |H_d^-|^2) \text{\indent}\text{(from soft breaking terms)}\nonumber\\
&& + [b(H_u^+H_d^- - H_u^0H_d^0) + \text{c.c.}]
\label{eq:MSSMscalarpot}
\end{eqnarray}
This potential has 8 d.o.f.\ from 4 complex scalar fields $H_u^+$, $H_u^0$, $H_d^0$ and $H_d^-$. At the same time it has 7 parameters: the two Standard Model gauge couplings $g$ and $g'$, the magnitude of the supersymmetric parameter $\mu$, and the three soft-breaking parameters $b$, $m_{H_u}^2$ and $m_{H_d}^2$. Notice how if $b=m_{H_u}^2=m_{H_d}^2=0$, meaning no supersymmetry breaking terms, the Higgs potential has a global minimum at $V=0$ for zero values of all the Higgs fields. In this case there is no EWSB, so EWSB is intimately connected to supersymmetry breaking in the MSSM.

@@

%%%%%%%%%%%%%%%%%
\section{Higgs boson properties}
%%%%%%%%%%%%%%%%%
Of the eight d.o.f.\ in the scalar potential for the Higgs component fields three are Goldstone bosons that get eaten by $Z$ and $W^\pm$ to give them masses. The remaining five d.o.f.\ form two neutral scalars $h$, $H$, two charged scalars $H^\pm$ and one neutral pseudo-scalar (CP-odd) $A$.\footnote{In addition to the scalars, we know that the Higgs supermultiplets contain four fermions, $\tilde{H}^0_u$, $\tilde{H}^0_d$, $\tilde{H}^+_u$ and $\tilde{H}^-_d$ (higgsinos). We will see later that these mix with the fermion partners of the gauge bosons (gauginos).} At tree level one can show that these have the masses:
\begin{eqnarray}
m_A^2 &=& \frac{2b}{\sin2\beta} = 2|\mu|^2 + m_{H_u}^2 + m_{H_d}^2,\\
m_{h, H}^2 &=& \frac{1}{2}\left(m_A^2 + m_Z^2 \mp \sqrt{(m_A^2- m_Z^2)^2 + 4m_Z^2m_A^2\sin^22\beta}\right),\\
m_{H^\pm}^2 &=& m_A^2 + m_W^2.
\end{eqnarray}
As a consequence $m_A$ and $\tan\beta$ can be used to parametrise the masses of the Higgs sector (at tree level), and
$H$, $H^\pm$ and $A$ are in principle unbounded in mass since they grow as $b/\sin2\beta$. However, at tree level the lightest Higgs boson is restricted to
\begin{equation}
m_h < m_Z|\cos 2\beta| < 91.2~\text{GeV}.
\end{equation}
In contrast we have the current best measurement of the Higgs boson mass of $m_h=125.10\pm 0.14$\,GeV, combining results from the LHC~\cite{ParticleDataGroup:2020ssz}.

Fortunately, there are large loop-corrections or the MSSM would have been excluded already.\footnote{It is worth pointing out here that the MSSM, despite its many parameters, is a falsifiable theory. For example, had the Higgs boson mass been $\sim15$\,GeV higher, which is perfectly allowed in the Standard Model, the MSSM would have been excluded.} Because of the size of the Yukawa couplings the largest corrections to the mass of the lightest Higgs comes from loops with top quarks and its supersymmetric partners, the scalar top quarks, or stops, $\tilde t_L$ and $\tilde t_R$. See Fig.~\ref{fig:hierarchy} for the relevant Feynman diagrams. In the limit where the mass of the stop quarks are larger than the top, $m_{\tilde{t}_R},m_{\tilde{t}_L}\gg m_t $, and with stop mass eigenstates close to the chiral eigenstates (more on this later), we get the dominant loop correction
\begin{equation}
\Delta m_h^2 = \frac{3}{4\pi^2} \cos^2\alpha\, y_t^2 m_t^2 \ln\left(\frac{m_{\tilde{t}_L}m_{\tilde{t}_R}}{m_t^2}\right),\label{eq:mhchiralstop}
\end{equation}
where $\alpha$ is a mixing angle for $h$ and $H$ with respect to the superfield component fields $H_u^0$ and $H_d^0$, given by
\begin{equation}
\frac{\sin\alpha}{\sin\beta}=-\frac{m_H^2+m_h^2}{m_H^2-m_h^2},
\end{equation}
at tree level.

@@

Since the lightest stop quark is expected to be the lightest squark in scenarios with universal soft masses at some high scale -- the reasoning here is the large downward RGE running of $m_{33}^Q$ from a common squark soft mass at some high scale due to the large top Yukawa coupling -- the expected sparticle spectrum lies mostly above 1 TeV, with the possible exception of gauginos/higgsinos. This points to so-called {\bf Split-SUSY} scenarios with heavy scalars and light gauginos, and a relatively large degree of fine-tuning. If one can live with this little hierarchy problem, it will explain why no signs of supersymemtry have been seen yet at the LHC.

@@

>>>>

Some  examples of other $\beta$-functions in the MSSM:
\begin{itemize}
\item For the supersymmetry parameter $\mu$
\begin{equation}
\beta_{\mu}\equiv \frac{d}{dt} \mu = \frac{1}{16\pi^2}\mu [3|y_t|^2+3|y_b|^2+|y_\tau|^2-3g_2^2-\frac{3}{5}g_1^2],
\label{eq:mu_beta}
\end{equation}
where $y_t$, $y_b$ and $y_\tau$, are the top and bottom quark, and tau lepton Yukawa couplings. Other Yukawa couplings also contribute to one-loop order, but due to their small size they can usually be safely ignored. Since the gauge couplings appear on the right-hand side here, we see that the differential equations that determine the running of $\mu$ and the gauge couplings are coupled. It is a general property of the $\beta$-functions that they couple the running of all parameters, which is caused by the (eventual) appearance of all particles as virtual particles in loops when we go to high-enough order.
\item For the soft-breaking parameters $M_i$ in (\ref{eq:gensoftterms}) we get the one-loop $\beta$-functions
\begin{equation}
\beta_{M_i}\equiv \frac{d}{dt}M_i = \frac{1}{8\pi^2}g_i^2M_ib_i,
\label{eq:beta_Mi}
\end{equation}
where $b_i$ is given above.
\item For the Higgs soft masses
\begin{eqnarray}
\beta_{m_{H_u}^2} &\equiv& \frac{dm_{H_u}^2}{dt} = \frac{1}{16\pi^2}(X_t-\frac{6}{5} g_1^2|M_1|^2- 6g_2^2|M_2|^2)  ,\label{eq:beta_Hu2}\\
\beta_{m_{H_d}^2} &\equiv& \frac{dm_{H_d}^2}{dt} = \frac{1}{16\pi^2}( X_b +X_\tau -\frac{6}{5} g_1^2|M_1|^2 - 6g_2^2|M_2|^2) \label{eq:beta_Hd2},
\end{eqnarray}
where we have used the standard abbreviations
\begin{eqnarray*}
X_t &\equiv& 6 |y_t|^2(m_{H_u}^2 + m_{Q_3}^2 + m_{u_3}^2) + 6|a_t|^2,\\
X_b &\equiv& 6 |y_b|^2(m_{H_d}^2 + m_{Q_3}^2 + m_{d_3}^2) + 6|a_b|^2,\\
X_\tau &\equiv& 2 |y_\tau|^2(m_{H_d}^2 + m_{L_3}^2 + m_{e_3}^2) + 2|a_\tau|^2,
\end{eqnarray*}
and where $m_{Q_3} = m_{33}^Q$, $m_{u_3}  = m_{33}^u$, $m_{d_3} = m_{33}^d$, $a_t=a_{33}^u$, $a_b=a_{33}^d$, and $a_\tau=a_{33}^e$,  in a simplification of our previous notation for the soft-term parameters in Sec.~\ref{sec:MSSM_soft_terms}.  
\end{itemize}

@@

For the scalar Higgs component fields in the MSSM this gives the potential
\begin{eqnarray}
%\label{eq:scaH}
V(H_u,H_d) &=& |\mu|^2 (|H_u^0|^2 + |H_u^+|^2 + |H_d^0|^2 + |H_d^-|^2)  \text{\indent \indent\indent \indent\indent\indent\indent}\text{(from $F$-terms)} \nonumber\\
&&+\frac{1}{8} (g^2 + g'{}^2)(|H^0_u|^2 + |H_u^+|^2 - |H^0_d|^2 - |H_d^-|^2)^2  \text{\indent\indent\indent\indent(from $D$-terms)} \nonumber\\
&&+ \frac{1}{2}g^2|H^+_uH_d^0{}^* + H^0_u H^-_d{}^*|^2 \nonumber\\
&&+ m_{H_u}^2(|H_u^0|^2 + |H_u^+|^2) + m_{H_d}^2(|H_d^0|^2 + |H_d^-|^2) \text{\indent}\text{(from soft breaking terms)}\nonumber\\
&& + [b(H_u^+H_d^- - H_u^0H_d^0) + \text{c.c.}]
\label{eq:MSSMscalarpot}
\end{eqnarray}
This potential has 8 d.o.f.\ from 4 complex scalar fields $H_u^+$, $H_u^0$, $H_d^0$ and $H_d^-$. At the same time it has 7 parameters: the two Standard Model gauge couplings $g$ and $g'$, the magnitude of the supersymmetric parameter $\mu$, and the three soft-breaking parameters $b$, $m_{H_u}^2$ and $m_{H_d}^2$. Notice how if $b=m_{H_u}^2=m_{H_d}^2=0$, meaning no supersymmetry breaking terms, the Higgs potential has a global minimum at $V=0$ for zero values of all the Higgs fields. In this case there is no EWSB, so EWSB is intimately connected to supersymmetry breaking in the MSSM.

@@

Another way to look at this is to solve (\ref{eq:EWSB_condition_Hd0_vevs}) and  (\ref{eq:EWSB_condition_Hu0_vevs})  for the well-measured $Z$-boson mass and $\sin(2\beta)$:
\begin{eqnarray*}
\sin(2\beta) &=& \frac{2b}{m_{H_d}^2+m_{H_u}^2+2|\mu|^2}, \\
M_Z^2 &=& \frac{|m_{H_d}^2-m_{H_u}^2|}{\sqrt{1-\sin^2(2\beta)}}-m_{H_u}^2-m_{H_d}^2-2|\mu|^2.
\end{eqnarray*}
Now, unless all the soft-masses $b$, $m_{H_u}^2$ and $m_{H_d}^2$ and the supersymmetric parameter $\mu$ is close to the electroweak scale there must be a great fine-tuning of parameters to get the correct $Z$-boson mass, and $\mu$ can certainly not be very much larger than the other three parameters.

%%%%%%%%%%%%%%%%%
\section{Higgs boson properties}
%%%%%%%%%%%%%%%%%
Of the eight d.o.f.\ in the scalar potential for the Higgs component fields three are Goldstone boson (real scalar fields) $G$ and $G^\pm$ that get eaten by $Z$ and $W^\pm$ to give them masses. The remaining five d.o.f.\ form two neutral (real) scalars $h$, $H$, two charged scalars $H^\pm$ and one neutral pseudo-scalar (CP-odd) $A$.\footnote{In addition to the scalars, we know that the Higgs supermultiplets contain four fermions, $\tilde{H}^0_u$, $\tilde{H}^0_d$, $\tilde{H}^+_u$ and $\tilde{H}^-_d$ (higgsinos). We will see later that these mix with the fermion partners of the gauge bosons (gauginos).}

We can decompose the complex Higgs fields in these fields by writing them as
\begin{eqnarray*}
\left( \begin{matrix} H_u^0 \\ H_d^0 \end{matrix}\right) &=& \left( \begin{matrix} v_u \\ v_d \end{matrix}\right) +\frac{1}{\sqrt{2}} R_\alpha \left( \begin{matrix} h \\ H \end{matrix}\right)  +\frac{i}{\sqrt{2}} R_\beta \left( \begin{matrix} G \\ A \end{matrix}\right),  \\
\left( \begin{matrix} H_u^+ \\ H_d^{-*} \end{matrix}\right) &=& R_\beta \left( \begin{matrix} G^+ \\ H^+ \end{matrix}\right),
\end{eqnarray*}
where the (orthogonal) mixing matrices are parameterised as
\begin{equation}
R_\alpha = \left( \begin{matrix} \cos\alpha &  \sin\alpha \\  -\sin\alpha & \cos\alpha \end{matrix}\right), \quad  R_\beta = \left( \begin{matrix} \sin\beta &  \cos\beta \\  -\cos\beta & \sin\beta \end{matrix}\right),
\end{equation}
Inserting these in to the quadratic terms in (\ref{eq:MSSMscalarpot})---the mass terms---we get the Lagrange terms
\begin{equation}
V=\half m_h^2 h^2 +\half m_H^2 H^2 +\half m_G^2 G^2 +\half m_A^2 A^2 + m^2_{G^\pm} |G^\pm|^2 + m^2_{H^\pm}
 |H^+|^2,
 \end{equation}
where the tree-level masses are
\begin{eqnarray}
m_G^2 &=& 0, \label{eq:HiggsmassesG} \\
m_{G^\pm}^2 &=& 0, \\
m_A^2 &=& \frac{2b}{\sin2\beta} = 2|\mu|^2 + m_{H_u}^2 + m_{H_d}^2,\\
m_{h, H}^2 &=& \frac{1}{2}\left(m_A^2 + m_Z^2 \mp \sqrt{(m_A^2- m_Z^2)^2 + 4m_Z^2m_A^2\sin^22\beta}\right),\\
m_{H^\pm}^2 &=& m_A^2 + m_W^2, \label{eq:HiggsmassesH}
\end{eqnarray}
with 
\begin{equation}
\frac{\sin\alpha}{\sin\beta}=-\frac{m_H^2+m_h^2}{m_H^2-m_h^2}.
\end{equation}

As a consequence $m_A$ and $\tan\beta$ can be used to parametrise the masses of the Higgs sector (at tree level), and
$H$, $H^\pm$ and $A$ are in principle unbounded in mass since they grow as $b/\sin2\beta$. However, at tree level the lightest Higgs boson is restricted to
\begin{equation}
m_h < m_Z|\cos 2\beta| < 91.2~\text{GeV}.
\end{equation}
In contrast we have the current best measurement of the Higgs boson mass of $m_h=125.10\pm 0.14$\,GeV, combining results from the LHC~\cite{ParticleDataGroup:2020ssz}.

Fortunately, there are large loop-corrections or the MSSM would have been excluded already.\footnote{It is worth pointing out here that the MSSM, despite its many parameters, is a falsifiable theory. For example, had the Higgs boson mass been $\sim15$\,GeV higher, which is perfectly allowed in the Standard Model, the MSSM would have been excluded.} Because of the size of the Yukawa couplings the largest corrections to the mass of the lightest Higgs comes from loops with top quarks and its supersymmetric partners, the scalar top quarks, or stops, $\tilde t_L$ and $\tilde t_R$. See Fig.~\ref{fig:hierarchy} for the relevant Feynman diagrams. In the limit where the mass of the stop quarks are larger than the top, $m_{\tilde{t}_R},m_{\tilde{t}_L}\gg m_t $, and with stop mass eigenstates close to the chiral eigenstates (more on this later), we get the dominant loop correction
\begin{equation}
\Delta m_h^2 = \frac{3}{4\pi^2} \cos^2\alpha\, y_t^2 m_t^2 \ln\left(\frac{m_{\tilde{t}_L}m_{\tilde{t}_R}}{m_t^2}\right).\label{eq:mhchiralstop}
\end{equation}

@@

Since the lightest stop quark is expected to be the lightest squark in scenarios with universal soft masses at some high scale---the reasoning here is the large downward RGE running of $m_{33}^Q$ from a common squark soft mass at some high scale due to the large top Yukawa coupling---the expected sparticle spectrum lies mostly above 1 TeV, with the possible exception of gauginos/higgsinos. This points to so-called {\bf Split-SUSY} scenarios with heavy scalars and light gauginos, and a relatively large degree of fine-tuning. If one can live with this little hierarchy problem, it will explain why no signs of supersymemtry have been seen yet at the LHC.

@@

\begin{Exercise}[]
Make the insertion into the scalar potential to show that we get the Higgs masses in (\ref{eq:HiggsmassesG})--(\ref{eq:HiggsmassesH}), and find the requirement on the mixing parameter $\alpha$.
\end{Exercise}
